---
title: "R Package Installations"
format: gfm
---

[![Install an R package](https://github.com/gvegayon/r-pkg-installs/actions/workflows/install_an_r_package.yaml/badge.svg)](https://github.com/gvegayon/r-pkg-installs/actions/workflows/install_an_r_package.yaml)

This repository contains a workflow that analyzes different strategies for installing R packages. Particularly, it compares using `install.packages()` vs `pak::pkg_install()` in different combinations of CRAN versions as well as architectures. The main workflow [here](./.github/workflows/install_an_r_package.yaml) runs the installs and saves the output as csv files under [data](./data).

## Preliminary analysis

```{r}
#| label: load-data
#| echo: false
#| message: false
#| warning: false
library(ggplot2)
library(data.table)
fns <- list.files("data", full.names = TRUE, pattern = "\\.csv$")

installation_data <- lapply(fns, fread) |>
  rbindlist(fill = TRUE)

# Failed installs have a string NA (yes, string)
failed_installs <- installation_data[installed_version == "NA"]
installation_data <- installation_data[installed_version != "NA"]

# Minor data wrangling
installation_data[, method := ifelse(
  use_pak, "pak::pkg_install()", "install.packages()"
  )]

# Combining fields
installation_data[, pak_source := sprintf(
  "%s (%s)",
  ifelse(use_pak, "pak::pkg_install()", "install.packages()"),
  installed_from
)]
```

As of `r Sys.Date()`, the following tables show the distribution of the successful installations by method and source of the installed package (total: `r nrow(installation_data)` successful installations vs `r nrow(failed_installs)` failed installations):

```{r}
#| label: summary-table
#| echo: false
#| message: false
#| warning: false
tab_pak <- installation_data[use_pak == TRUE, table(cran, installed_from)]
tab_no_pak <- installation_data[use_pak == FALSE, table(cran, installed_from)]

knitr::kable(tab_pak, caption = "Using pak::pkg_install()")
knitr::kable(tab_no_pak, caption = "Using install.packages()")
```

The following figure shows the installation times (in seconds) for the `data.table` package using different methods and CRAN versions. So far, `pak` is only using source for installations, even when the CRAN repository is set to the specific OS architecture (e.g., `__linux__/jammy`):

```{r}
#| label: speed
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Installation times (in seconds) for the `data.table` package using different methods and CRAN versions."
ggplot(installation_data, aes(x = install_time)) +
  geom_histogram(aes(fill = method), position = "dodge") +
  theme_minimal() +
  labs(
    x = "Package",
    y = "Time (seconds)",
    fill = "Installation Method",
    caption = sprintf(
      "Data as of %s, including %d installations",
      Sys.Date(),
      nrow(installation_data)
      )
  ) +
  facet_wrap(~ cran) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Looking at using source vs binary installation:

```{r}
#| label: source-vs-binary
#| echo: false
#| message: false
#| warning: false
ggplot(installation_data, aes(x = install_time)) +
  geom_histogram(aes(fill = pak_source), position = "dodge") +
  theme_minimal() +
  labs(
    x = "Package",
    y = "Time (seconds)",
    fill = "Installation type",
    caption = sprintf(
      "Data as of %s, including %d installations",
      Sys.Date(),
      nrow(installation_data)
      )
  ) +
  facet_wrap(~ cran) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Implementation details

### The installation process

1. Initialize the job inside the corresponding rocker container.
2. Try to install the `data.table` package (needs compilation).
3. Print the installation output. Using that, check whether a source or a binary version was installed (check the regex `installing [*]source[*]` if it is `install.packages` or `Building {}` if it is `pak::pkg_install`).
4. Record the installation time and any relevant system information. The filename is a hash of the file itself.
5. The output is stored as an artifact. Once all jobs are complete, the artifacts are combined into a single CSV file, which is then uploaded to the repository via a commit.

## Random observations

- [2025-09-10 gvegayon] `pak` does not seem to be working as expected. All instalations using the `pak` package in the `rocker/r-ver` containers have been from source, even when the CRAN repository is set to the specific OS architecture (e.g., `__linux__/jammy`). 

- [2025-09-03 gvegayon] The posit public package manager (p3m) seems to contain multiple snapshot versions of compiled packages. As of September 3, 2025, the latest 8 versions of data.table (from 1.16.0 in October 10, 2024 up to 1.17.8 in July 10, 2025) are available as binaries for Mac OS arm64 chips. Nonetheless, running the same installer in the `rocker/r-ver:4.5.1` container results in a different behavior, always installing the source version of the package.